---
title: "Ch 1 Larval CADO w/ multiple populations"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE)
```

# Setup
```{r include=TRUE}
library(ggplot2)
library(tidyr)
library(plyr)
library(dplyr)
library(qvalue)
library(stringr)
library(pcadapt)
library(poolSeq)
library(ACER)
library(ggman)
library(VennDiagram)
library(scales)
library(data.table)
library(patchwork)
```

```{r include=TRUE}
# Function to convert p-values to q-values
Qvalue_convert <- function(table) {


table <- spread(table,GROUP,PVAL)  
  
table$CHR <- table$CHROM

table %>% 
  mutate(CHR = str_replace(CHR, "NC_035780.1", "1")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035781.1", "2")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035782.1", "3")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035783.1", "4")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035784.1", "5")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035785.1", "6")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035786.1", "7")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035787.1", "8")) %>%
  mutate(CHR = str_replace(CHR, "NC_035788.1", "9")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035789.1", "10"))  -> pp1

pp1 <- unite(pp1, "SNP", c("CHROM","BP"),remove = FALSE)

pp1$CHR <- as.numeric(pp1$CHR)  

# Set pi0=1 to equal Benjamini Hochberg
pp1$QSTR <- qvalue(pp1$PSTR, pi0 = 1)$qvalues
pp1$QCON <- qvalue(pp1$PCON, pi0 = 1)$qvalues


return(pp1)
}
```

```{r include=TRUE}
# Function to determine significant loci in various outputs
Significant_subset <- function(pv, alpha, alpha2) {
  # Add new logical columns based on conditions
  pv$Sig.STR <- pv$QSTR < alpha
  pv$Sig.CON <- pv$QCON < alpha
  
  # Subset the dataframe
  ppsig <- subset(pv, QSTR < alpha )
  ppsig <- subset(ppsig, QCON > alpha2)
  
  # Print diagnostics
  print(nrow(pv))
  print(nrow(ppsig))
  print(nrow(ppsig) / nrow(pv))
  
  # Return the modified dataframe
  return(ppsig)
}
```

```{r include=TRUE}
group_and_average <- function(df) {
  df %>%
    dplyr::group_by(SNP) %>%
    dplyr::summarize(
      across(
        c(PCON, PSTR, CHR, QCON, QSTR),
        ~ if (all(is.na(.))) NA else min(., na.rm = TRUE)
      ),
      Sig.STR = any(Sig.STR, na.rm = TRUE),
      CHROM  = first(CHROM),
      BP     = first(BP),
      .groups = "drop"
    )
}
```

```{r include=TRUE}
cbPaletteSmall <- c("purple", "green", "gold", "red", "orange", "blue", "violet")
```

```{r}
sessionInfo()
```


# Visualize across 10,000 random loci

## All
```{bash, eval = FALSE}
mamba activate CADO

cut -f1,2 Ch1.dp25.cov.sync | sort -u  > filtered.loci

mawk '!/CHR/' filtered.loci | shuf -n 50000  | shuf -n 10000 > rand.loci

mawk 'NR==FNR{a[$1,$2]; next} ($1,$2) in a' rand.loci Ch1.dp25.cov.sync | cut --complement -f67- > input.all.rand.sync

```

```{r}
input.all.rand <- read.sync(file="input.all.rand.sync", gen=seq(1,63), repl=seq(1,63), polarization="rising")


input.all.rand.af <- af(input.all.rand,gen=seq(1,63), repl=seq(1,63))

bio.labels <- c(
  "ARC2_CB1","ARC2_CB2","ARC2_CB3","ARC2_SB1","ARC2_SB2","ARC2_SB3","ARC2_T01","ARC2_T02","ARC2_T03",
  "MV1_CB1","MV1_CB2","MV1_CB3","MV1_SB1","MV1_SB2","MV1_SB3","MV1_T01","MV1_T02","MV1_T03",
  "MV2_CB1","MV2_CB2","MV2_CB3","MV2_SB1","MV2_SB2","MV2_SB3","MV2_T01","MV2_T02","MV2_T03",
  "MV3_CB1","MV3_CB2","MV3_CB3","MV3_SB1","MV3_SB2","MV3_SB3","MV3_T01","MV3_T02","MV3_T03",
  "NEH1_CB1","NEH1_CB2","NEH1_CB3","NEH1_SB1","NEH1_SB2","NEH1_SB3","NEH1_T01","NEH1_T02","NEH1_T03",
  "NEH2_CB1","NEH2_CB2","NEH2_CB3","NEH2_SB1","NEH2_SB2","NEH2_SB3","NEH2_T01","NEH2_T02","NEH2_T03",
  "NEH3_CB1","NEH3_CB2","NEH3_CB3","NEH3_SB1","NEH3_SB2","NEH3_SB3","NEH3_T01","NEH3_T02","NEH3_T03"
)
length(bio.labels)  # should be 63

pop.ids <- colnames(input.all.rand.af)  # "F1.R1.freq", ..., "F63.R63.freq"

meta <- data.frame(
  sample_id   = pop.ids,
  bio_label   = bio.labels,
  stringsAsFactors = FALSE
)

# Derive treatment from the biological label
meta$treatment <- ifelse(grepl("_CB", meta$bio_label), "control",
                  ifelse(grepl("_SB", meta$bio_label), "stress", "initial"))
# can also parse by trial  
meta$trial <- sub("_.*", "", meta$bio_label)
# and parse by population
meta$pop <- ifelse(grepl("^ARC", meta$bio_label), "ARC",
                   ifelse(grepl("^MV",  meta$bio_label),  "MV",
                          ifelse(grepl("^NEH", meta$bio_label), "NEH", NA)))
View(meta)

#check that column names in input.all.rand.af and sample_id column in meta table are the same
all.equal(colnames(input.all.rand.af), meta$sample_id)
# should be TRUE

# assign the columns of input.all.rand.af with their proper sample names instead of the sample_id created by inputting the table to poolseq
colnames(input.all.rand.af) <- meta$bio_label
```


### PCA of random 10K loci from all samples 
shows a strong pattern of population and experiment year (2023 – ARC2, NEH1, MV1, MV2; 2024 – NEH2, NEH3, MV3)
```{r}
# Perform PCA
af_t <- t(input.all.rand.af)

pca_res <- prcomp(af_t, center = TRUE, scale. = TRUE)

# Get scores for plotting
pca_df <- as.data.frame(pca_res$x)
#pca_df$locus <- rownames(input.all.rand.af)  # keep locus IDs if you have them

var_explained <- pca_res$sdev^2 / sum(pca_res$sdev^2) * 100

pca_df$Treatment <- meta$treatment
pca_df$Trial <- meta$trial

pca_rand10k <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Trial, fill=Trial)) +
  geom_point(aes(alpha=0.2), size =4,shape=21, col="black")+  scale_fill_manual(values=cbPaletteSmall,name="Exp. Trial")+ 
guides(alpha=FALSE) +theme(axis.title.x = element_text(size = 24),axis.title.y = element_text(size = 24)) +
  theme_minimal() +
  labs(
    x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
    y = paste0("PC2 (", round(var_explained[2], 1), "%)"),
    title = "PCA of Allele Frequencies"
  ); pca_rand10k

ggsave(filename = "PCA/ch1_pca_rand10k.png", plot = pca_rand10k, width = 6, height = 5, units = "in", dpi = 300)
```

## PCA of random 10K loci from All Initial Samples
As expected, shows the same strong pattern of population and experiment year seen in the all sample plot
```{r}
#subset data to only T0 initial samples
t0_cols <- grepl("_T0", colnames(input.all.rand.af))
input.T0.rand.af <- input.all.rand.af[, t0_cols]
afi_t <- t(input.T0.rand.af)

#subset metadata table
t0_meta <- subset(meta, meta$treatment == "initial")
View(t0_meta)

# start pca code
pca_i_res <- prcomp(afi_t, center = TRUE, scale. = TRUE)

# Get scores for plotting
pca_i_df <- as.data.frame(pca_i_res$x)

var_explained <- pca_i_res$sdev^2 / sum(pca_i_res$sdev^2) * 100

pca_i_df$Treatment <- t0_meta$treatment
pca_i_df$Trial <- t0_meta$trial

pca_t0rand10k<- ggplot(pca_i_df, aes(x = PC1, y = PC2, color = Trial, fill=Trial)) +
  geom_point(aes(alpha=0.2), size =10,shape=21, col="black")+  scale_fill_manual(values=cbPaletteSmall,name="Exp. Trial")+ 
guides(alpha=FALSE) +theme(axis.title.x = element_text(size = 24),axis.title.y = element_text(size = 24)) +
  theme_classic() +
  labs(
    x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
    y = paste0("PC2 (", round(var_explained[2], 1), "%)"),
    title = "PCA of Allele Frequencies"
  );pca_t0rand10k

ggsave(filename = "PCA/ch1_pca_initial_rand10k.png", plot = pca_t0rand10k, width = 6, height = 5, units = "in", dpi = 300)
```


## Random loci for each individual experimental population (ARC, NEH, MV)


Random loci for each experimental block - Jon's code 
```{r}
# Ensure meta is keyed by bio_label and in column order
meta <- meta[match(colnames(input.all.rand.af), meta$bio_label), ]
stopifnot(all(meta$bio_label == colnames(input.all.rand.af)))

pops <- c("ARC", "MV", "NEH")

for (pop in pops) {
  # select columns belonging to this population
  cols_idx <- which(meta$pop == pop)
  cat(sprintf("\n== Population %s: found %d sample columns ==\n", pop, length(cols_idx)))
  if (length(cols_idx) < 2) {
    warning(sprintf("Population %s: less than 2 samples found - skipping PCA.", pop))
    next
  }

  sub_mat <- input.all.rand.af[, cols_idx, drop = FALSE]
  cat("  dim(sub_mat) (loci x samples):", dim(sub_mat), "\n")

  # transpose: rows = samples, cols = loci
  mat_t <- t(sub_mat)

  # drop zero-variance loci within this population
  var_cols <- apply(mat_t, 2, var, na.rm = TRUE)
  keep     <- which(var_cols > 0)
  removed_n <- sum(var_cols == 0)
  cat("  loci removed (zero variance):", removed_n, "\n")

  if (length(keep) < 2) {
    warning(sprintf("Population %s: fewer than 2 variable loci after filtering - skipping PCA.", pop))
    next
  }
  mat_t2 <- mat_t[, keep, drop = FALSE]
  cat("  dim(mat_t2) (samples x retained loci):", dim(mat_t2), "\n")

  # PCA
  pca_res <- prcomp(mat_t2, center = TRUE, scale. = TRUE)
  pca_df  <- as.data.frame(pca_res$x)

  # add metadata for these samples
  sample_labels <- rownames(mat_t2)                      # bio_labels
  sample_meta   <- meta[match(sample_labels, meta$bio_label), ]

  pca_df$population <- sample_meta$population            # should all be 'pop'
  pca_df$treatment  <- sample_meta$treatment             # T0 / CB / SB
  pca_df$bio_label  <- sample_meta$bio_label
  pca_df$trial <- sample_meta$trial

  var_explained <- (pca_res$sdev^2) / sum(pca_res$sdev^2) * 100

  # palette for treatment or population; here color by treatment
  if (exists("cbPaletteSmall")) {
    fill_scale <- scale_fill_manual(values = cbPaletteSmall, name = "Treatment")
  } else {
    fill_scale <- scale_fill_brewer(palette = "Set1", name = "Treatment")
  }

  # plot: within a population, color by treatment
  p <- ggplot(pca_df, aes(x = PC1, y = PC2, fill = treatment)) +
    geom_point(size = 5, shape = 21, color = "black", alpha = 0.8) +
    fill_scale +
    geom_text_repel(aes(label = trial), size = 3) +
    theme_classic() +
    theme(axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          plot.title   = element_text(hjust = 0.5)) +
    labs(
      title = paste0("PCA - Population ", pop),
      x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
      y = paste0("PC2 (", round(var_explained[2], 1), "%)")
    )

  outfile <- paste0("PCA_population_", pop, ".png")
  ggsave(outfile, plot = p, width = 10, height = 7, dpi = 300)
  cat("  saved:", outfile, "\n")
}

```
